Wind veer and speed in turbulent Ekman flow part I: scaling analysis and velocity profile model

### 2025-03-27 write1/REVISION1/manuscript.tex
finished work on Reviewer 1 responses
finished work on Review two and most part of no. 3

% additional packages
\usepackage{xcolor}
\usepackage{colortbl}
\usepackage{booktabs} 
\usepackage{amssymb}
\usepackage{hyperref}
\setcitestyle{aysep={}}
\linenumbers

@@

\date{\scriptsize Submitted \today}
\title{Wind veer and speed in turbulent Ekman flow part I: scaling analysis and universal profile model}
\titlerunning{A Universal Velocity Profile for Ekman Flow}
\author{Cedrick Ansorge$^1$ and Hauke Wurps$^{2,3}$}
\authorrunning{C Ansorge and H Wurps}
\institute{$^{1}$\,FU Berlin, Institut f\"ur Meteorologie, Carl-Heinrich-Becker-Weg 6--10, 12165 Berlin, Germany {cedrick@posteo.de}\\
  $^{2}$\,Carl von Ossietzky Universit\"at Oldenburg, School of Mathematics and Science, Institute of Physics \\ 
  $^{3}$\,ForWind - Center for Wind Energy Research, K\"upkersweg 70, 26129 Oldenburg, Germany 
  }

\begin{document} 
%
\maketitle
%
\begin{abstract}
  The profiles of wind speed and direction in turbulent Ekman flow
  are formulated based on asymptotic theory and data from direct numerical simulation. 
  %
  The profile of the streamwise component follows the classical viscous, logarithmic and wake scaling.
  %
  In the outer layer, the velocity component profiles can be described by an Ekman-spiral with adapted
  boundary conditions that result in a reduction of the spiral-like rotation. 
  %
  The span-wise component poses a conceptual challenge to the channel-flow analogy
  in the context of asymptotic matching; it exhibits a mixed scaling in the surface layer, but follows
  outer scaling for most of the outer layer.
  %
  Viscous stress scales universally across the boundary layer in inner units while the total stress
  becomes universal as a function of outer height.
  %
  This implies a mixed scaling for the turbulent stress and eddy viscosity across the inner layer
  and convergence to a universal scaling as function of the outer height across the outer layer
  for increasing scale separation vide Reynolds numbers. 
  % 
  The extrapolation to these scaling to atmospheric scale separation is confirmed via large-eddy simulation 
  in part II of this manuscript. 
\end{abstract}
~\\[1em] 
\textbf{Keywords} Direct Numerical Simulation $\cdot$ Scale Separation $\cdot$ Ekman Layer $\cdot$ Surface Layer $\cdot$ Hodograph

@@

The Coriolis force bends the apparent path of motion on a rotating sphere and establishes geostrophic equilibrium when in balance with a pressure gradient force. Wind veer away from the wind direction in geostrophic equilibrium is
(i) due to direct frictional effects in the very vicinity of the surface and
(ii) due to turbulence which exerts indirect frictional effects; these effects
cause a slow-down of the mean wind reducing the Coriolis force thus turning the wind
in favor of the pressure gradient force. 
%%
Not only does the veering set the frame of reference for surface layer theory, it also has
effects at small and large scales from large-scale dispersion via plume spreading to
cyclone spin-down \citep{svensson:BM2009}
and on the capabilities of data assimilation and accuracy of surface flux estimates \citep{brown:QJR2005}. 
%
From a large-scale perspective, the veering of wind across the planetary boundary layer determines
the amount of cross-isobaric mass-flux, commonly referred to as 'Ekman pumping' \citep{ekman:AMA1905},
and it is thus a key factor in the life-cycle of large-scale synoptic systems.
%
% %% This shear is probably not related to Ekman-turning 
% On the meso-scale, shear in the surface layer a crucial factor for the initiation and sustained strong convective events (REFERENCE?).
% 
Within the atmospheric boundary layer (ABL), directional shear of the wind in
the upper part of the surface layer may cause a systematic yaw for tall wind power generation devices
where blades reach into the Ekman layer, i.e. that part of the boundary layer where the wind starts to turn;
an exact estimate of such effects is critical in the site assessments for wind farms
\citep{calaf:PF2010, mirocha:WES2018}. 
%
\par
%
In the planetary boundary layer, wind veer is characterized by the surface veering angle $\alpha$
defined as the angle between the negative surface shear stress $\tau_\SFC$ and the geostrophic wind.
%
Surface veering $\alpha$ and geostrophic drag $Z\equiv u_\star/G$, where the friction velocity $u_\star\equiv\sqrt{|\tau_\SFC|/\rho}$,
uniquely determine the surface drag $\tau_\SFC$ in a turbulent Ekman flow. 
%
In any quantitative description of the surface layer, the friction velocity $u_\star$ is the dynamic scale
and $\alpha$ defines the horizontal alignment of the frame of reference, i.e. the rotation of the surface 
friction with respect to the outer wind direction. 
%
Knowledge about $u_\star$ and $\alpha$ is thus a prerequisite for any quantitative theory of the surface layer, 
and \citet{rossby:PIP1935} constrained the two parameters based on integral relations in the ABL.
%
Asymptotic similarity theory was later used by \cite{tennekes:JAS1973, blackadar:JAS1968},
and--based on his seminal direct numerical simulations (DNS) of Ekman flow--,~\cite{spalart:JFM1989} suggested
a modification to take into account effects of low to intermediate Reynolds numbers.
Later on, constants were re-evaluated with a focus on the ABL based on
observations \citep{hogstrom:BM1988,hogstrom:BM1996}
and numerical modelling \citep{spalart:PF2008, spalart:PF2009, ansorge:BM2014, ansorge:BM2019}.
%
\par
%
Attempts were also undertaken to obtain profiles of the wind speed: 
One approach is to match the inner and outer layer at a reference height as suggested by
\cite{etling:2002, emeis:2018}  (Sec. 21.10; Eq. 21.48); they choose the Prandtl-layer height $z_\text{Prandtl}$ to match
the wind speed profiles, which, however, requires external prescription of $\alpha(z_\text{prandtl})$, the veering at that height. 
A one-dimensional profile with constant veering is given by \citet[Sec. 3; Eq. 3.1-3.19]{emeis:m2007}. 
%%%

@@

homogeneous, stationary ABL over a flat surface under neutral stratification. 
Universal profiles of the wind vector for turbulent Ekman flow not only are a well-described limit
for theoretical exploration or higher-order approaches taking into account possible effects of stratification,

@@

(iv) Barotropic and baroclinic effects on the velocity profile require to consider 

@@

The f-plane approximation is applied such that rotation only acts on horizontal velocity components;
we thus neglect rotational effects on the horizontal components of velocity and dynamical effects
due to latitudinal variation of the rate of rotation.

@@

correspond to an Ensemble average. 

@@

  \underline{\tau}_\mathrm{sfc} = \left(\begin{array}{c} \tau_x\\ \tau_y\\\tau_z\end{array}\right) = - \nu \left( \frac{\partial U}{\partial z} \hat{e}_x + \frac{\partial V}{\partial z} \hat{e}_y\right)

@@

  u_\star = \sqrt{ || - \underline{\tau}_\mathrm{sfc} || } 

@@

the only non-dimensional parameter that governs the system.

@@

@@

The Eddy viscosity plays a crucial part when modelling profiles and the vertical

@@

For the symmetries in the flow (horizontal homogeneity, and $W=0$), it is

@@

    Buffer layer (i=10),
    logarithmic Layer (i=100), and
    outer layer (i=400) of the case with $Re_\tau=2978$;

@@

\section{A universal velocity profile for the turbulent Ekman layer}

@@

boundary conditions at the surface but at a different height. 

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Implications for surface-layer scaling} 
Eq.~(\ref{eqn:scaling_spanwise}) establishes a universal mixed scaling for the spanwise velocity in the viscous layer:
%
While it requires the vertical coordinate to be expressed in inner units, the velocity itself is normalized by the geostrophic wind, and becomes inversely proportional to the friction Reynolds number $\RE_\tau=\delta^+$ when considered at a fixed height.
%
In vicinity of the surface, such mixed scaling has already been identified for higher-order statistics in convective flows \citep{mellado:BM2016,li:JAS2018}, where large scales leave their signatures in vicinity of the surface.
%
It is important to note here that, while $V$ is a first-order statistic from a statistical perspective, the spanwise velocity is a higher-order correction term from the perspective of similarity theory and from the viewpoint of the channel-flow analogy that is routinely employed in the surface layer.
%
Further, this is consistent with the scaling for the velocity hodograph found in Eq.~(\ref{eqn:hodograph_scaling}) where the friction velocity also drops out.
%
\par
%
In the surface layer, there is not only a mixed scaling--as we had already identified in the viscous layer--, but we cannot find a universal function onto which the profiles of spanwise velocity collapse. 
%
This additional degree of freedom reflects the inner--outer matching problem for the spanwise velocity. 
%
Rather than giving a universal profile for this region, we resort here to a parametric description of the problem in terms of the function $f_{V,\mathrm{log}}$ determined by the parameters $a_\mathrm{log}$, $b_\mathrm{log}$, $c_\mathrm{log}$ which can be estimated based on the above scaling considerations for any Reynolds number.
%
We note that, once the parameter $a_\mathrm{log}$ is known, the parameters $b_\mathrm{log}$ and $c_\mathrm{log}$ can be estimated solely based on $f_{V,\mathrm{visc}}$, i.e. using the value $v_{10}$ and $d_{10}$ found for the viscous region of the flow.
%
For the range of Reynolds number relevant to geophysical problems ($10^4\lesssim \RE_D \lesssim 10^6$), the variation of $c_\mathrm{log}$ is, however, rather small.

@@

%
%\begin{figure}
%  \includegraphics[width=\textwidth]{mag_dir.eps} 
%  \caption{Total velocity and veering (in degrees) vs inner and outer height.
%    Dashed lines show DNS data, thick lines are from semi-empirical theory.} 
%\end{figure} 
%
\section{Conclusions}
%
We investigate the wind veer and fundamental scaling properties of the velocity profiles in Ekman flow. 
%
Based on scaling considerations and direct numerical simulation spanning on decade in external separation from $Re_\Lambda=1.25 \times 10^5$ to $Re_\Lambda=1.28\times 10^6$, we derive a universal formulation for the both horizontal components of the velocity profile in a stationary, smooth Ekman layer.  
% 
This formulation is consistent with the DNS data and also yields reasonable results  at geophysical scale separation; the logarithmic law of the mean velocity is recovered with the well-known limits and deviations towards the surface and Ekman layer. 
%
The classic formulation of the Ekman layer, employing the surface boundary condition, is replaced by a modified solution that can be obtained by Ekman's system of governing equations, but using different boundary conditions that are more appropriate of the actual situation encountered in the planetary boundary layer. 
% 
The three parameters that characterize this boundary condition are estimated based on DNS data. 
% 
\par
% 
To quantify the spanwise velocity component consistently across the boundary layer, we derive a universal scaling of the spanwise velocity component in a shear-aligned reference frame. For the \emph{inner layer}, we find the mixed scaling  
%
\begin{equation} 
  V^{\alpha_\star}/G = \frac{1}{\delta^+} f(z^+)
\end{equation}
%
that is, the spanwise velocity normalized by the outer velocity scale is a universal function of the inner height and Friction Reynolds number $Re_\tau = \delta^+$. 
% 
This scaling is derived here based on scaling considerations, and it is in excellent agreement with the DNS data available.   
% 
In the outer layer, the spanwise velocity follows outer scaling, consistently with the Ekman model discussed above (Sec.~\ref{sec:ekman}).
% 
\par
% 
Our results suggest that there is no lower limit of the turning. 
% 
Hence---despite its very large scale separation / huge Reynolds number---the ABL is not 
in the 'limit' of high Reynolds number from the perspective of wind veer, but always in a regime where 
changes in $\RE$ impact on the vertical partitioning of rotation.

@@

\appendix
\section{Appendices}
\subsection{Laminar Ekman solution with consideration of inner layer}
\label{app:ekman_solution}
The following Ekman system is obtained by averaging the Navier--Stokes equations horizontally and over time and neglecting the turbulent transport terms (turbulence can then be considered via the eddy-viscosity concept through variations in the viscosity $\nu$):
\begin{subequations}
\begin{eqnarray}
  \left(\begin{matrix}
    \p_t U\\
    \p_t V  
  \end{matrix}\right)&=&\left(\begin{matrix}
     fV &+& \nu \p_z^2 U\\ 
    -f(U-G) &+& \nu \p_z^2 V
  \end{matrix}\right)\\ 
  \Rightarrow \partial_t (U+iV) &=& f(V-i(U-G))) + \nu \p_z^2(U+iV)
\end{eqnarray}
In stationary conditions, this system is solved by
\begin{eqnarray}
  \hat{u}(z) &=& U_{\infty} + e^{-\gamma z} \left[A \cos\gamma z + B \sin\gamma z\right]   \\
  \hat{v}(z) &=& V_{\infty} + e^{-\gamma z} \left[-A \sin\gamma z + B \cos\gamma z\right]
\end{eqnarray} 
\end{subequations}
where the constants $U_\infty$, $V_\infty$ set the top boundary condition and $A$ and $B$ set the bottom boundary condition. 
%
The most common boundary condition for a surface Ekman layer is $A=U_{\infty}=G$, $B=0$, and $V_{\infty}=0$.
%
The lower boundary condition, however, neglects the existence of the surface layer, and it appears reasonable to define
$A=c G$ where $c<1$ is a constant that incorporates the increased shear in the surface layer.
%
Given a 'matching height' $z_{match}$ and normalized matching height $\xi=\gamma z_\text{match}$ in the upper part of the inner layer, we can match the Ekman profile
to the inner layer by letting
%
\begin{subequations}
\begin{eqnarray}
  \begin{matrix} 
    u(z_\text{match}) \equiv u_\text{match} &=& U_{\infty} + e^{-\gzm} \left [ A \cos\gzm + B\sin\gzm \right] \\
    v(z_\text{match}) \equiv v_\text{match} &=& V_{\infty} + e^{-\gzm} \left [-A \sin\gzm + B\cos\gzm \right]
  \end{matrix} 
\end{eqnarray}
\begin{eqnarray}
  \Rightarrow \left(\begin{matrix}
    u_\text{match}-U_{\infty} \\ 
    v_\text{match}-V_{\infty}
  \end{matrix}\right) = e^{-\gzm}\left(\begin{matrix}A \\ B \end{matrix}\right)\left(\begin{matrix}
    \cos\gzm &+ \sin\gzm \\ -\sin\gzm &+ \cos\gzm
  \end{matrix} \right) \\
\end{eqnarray}
\end{subequations}
Matching the profile at $\xi=0$, one obtains
$A = \Delta u_\text{match} $ and $B=-\Delta w_\text{match} $; and when the direction $Ox$ is aligned
with the geostrophic wind, we obtain the textbook-case $A=|\mathbf{G}|$ and $B=0$.

@@

\vspace{1em}
\noindent\textbf{Acknowledgements} Simulations were carried out on the German supercomputing systems 
\texttt{hawk} at H\"ocstleistungsrechenzentrum Stuttgart (Bundesprojekt 44187) and \texttt{juwels} at Gauss Centre for supercomputing, 
For\-schungs\-zen\-trum J\"ulich (Projects \texttt{hku24} and \texttt{stadit}). 
We extend our thanks to Sally Issa for valuable feedback on the manuscript.

@@

>>>>

% additional packages
\usepackage{xcolor}
\usepackage{colortbl}
\usepackage{booktabs} 
\usepackage{amssymb}
\usepackage{hyperref}
\journalname{Bounday-Layer Meteorology}
\setcitestyle{aysep={}}
\linenumbers

@@

\date{\scriptsize Submitted \today}
\title{Wind veer and speed in turbulent Ekman flow part I: scaling analysis and velocity profile model}
\titlerunning{Scaling Analysis and Velocity Profile for Ekman Flow}
\author{Cedrick Ansorge$^1$ and Hauke Wurps$^{2,3}$}
\authorrunning{C Ansorge and H Wurps}
\institute{$^{1}$\,FU Berlin, Institut f\"ur Meteorologie, Carl-Heinrich-Becker-Weg 6--10, 12165 Berlin, Germany {cedrick@posteo.de}\\
  $^{2}$\,Carl von Ossietzky Universit\"at Oldenburg, School of Mathematics and Science, Institute of Physics \\ 
  $^{3}$\,ForWind - Center for Wind Energy Research, K\"upkersweg 70, 26129 Oldenburg, Germany 
  }

\begin{document} 
%
\maketitle
%
\begin{abstract}
  The profiles of wind speed and direction in turbulent Ekman flow
  are formulated based on asymptotic theory and data from direct numerical simulation. 
  %
  The profile of the streamwise component, considered in a wall-stress-aligned reference frame, follows the classical viscous, logarithmic and wake scaling.
  %
  In the outer layer, the velocity component profiles can be described by an Ekman-spiral with adapted
  boundary conditions that result in a reduction of the spiral-like rotation. 
  %
  The span-wise component poses a conceptual challenge to the channel-flow analogy
  in the context of asymptotic matching; it exhibits a mixed scaling in the surface layer, but follows
  outer scaling for most of the outer layer.
  %
  Viscous stress scales universally across the boundary layer in inner units while the total stress
  becomes universal as a function of the outer height, commonly denoted as $z^-$.
  %
  This implies a mixed scaling for the turbulent stress and eddy viscosity across the inner layer
  and convergence to a scaling as function of the outer height across the outer layer
  for increasing scale separation, i.e. for increasing Reynolds number. 
  % 
  The extrapolation to these scaling to atmospheric scale separation is confirmed via large-eddy simulation 
  in part II of this manuscript. 
\end{abstract}
~\\[1em] 
\textbf{Keywords} Direct Numerical Simulation $\cdot$ Scale Separation $\cdot$ Ekman Layer $\cdot$ Surface Layer $\cdot$ Hodograph

@@

The Coriolis force bends the apparent path of motion on a rotating sphere and establishes geostrophic equilibrium when in balance with a pressure gradient force. Wind veer away from the wind direction in geostrophic equilibrium is
(i) due to direct frictional effects in the very vicinity of the surface and
(ii) due to turbulence which exerts indirect frictional effects; these effects
cause a slow-down of the mean wind reducing the Coriolis force thus turning the wind
in favor of the pressure gradient force. 
%%
Not only does the veering set the frame of reference for surface layer theory, it also has
effects at small and large scales from large-scale dispersion via plume spreading to cyclone 
spin-down \citep{svensson:BM2009} and on the capabilities of data assimilation 
and accuracy of surface flux estimates \citep{brown:QJR2005}. 
%
From a large-scale perspective, the veering of wind across the planetary boundary layer determines
the amount of cross-isobaric mass-flux, commonly referred to as 'Ekman pumping' \citep{ekman:AMA1905},
and it is thus a key factor in the life-cycle of large-scale synoptic systems.
%
% %% This shear is probably not related to Ekman-turning 
% On the meso-scale, shear in the surface layer a crucial factor for the initiation and sustained strong convective events (REFERENCE?).
% 
Within the atmospheric boundary layer (ABL), directional shear of the wind in
the upper part of the surface layer may cause a systematic yaw for tall wind power generation devices
where blades reach into the Ekman layer, i.e. that part of the boundary layer where the wind starts to turn;
an exact estimate of such effects is critical in the site assessments for wind farms
\citep{calaf:PF2010, mirocha:WES2018}. 
%
\par
%
In the planetary boundary layer, wind veer is characterized by the surface veering angle $\alpha$
defined as the angle between the negative surface shear stress $\tau_\SFC$ and the geostrophic wind.
%
Surface veering $\alpha$ and geostrophic drag $Z\equiv u_\star/G$, where the friction velocity $u_\star\equiv\sqrt{|\tau_\SFC|/\rho}$,
uniquely determine the surface drag $\tau_\SFC$ in a turbulent Ekman flow. 
%
In any quantitative description of the surface layer, the friction velocity $u_\star$ is the dynamic scale
and $\alpha$ defines the horizontal alignment of the frame of reference, i.e. the rotation of the surface 
friction with respect to the outer wind direction. 
%
Knowledge about $u_\star$ and $\alpha$ is thus a prerequisite for any quantitative theory of the surface layer, 
and \citet{rossby:PIP1935} constrained the two parameters based on integral relations in the ABL.
%
Asymptotic similarity theory was later used by \cite{tennekes:JAS1973, blackadar:JAS1968},
and--based on his seminal direct numerical simulations (DNS) of Ekman flow--,~\cite{spalart:JFM1989} suggested
a modification to take into account effects of low to intermediate Reynolds numbers.
Later on, constants were re-evaluated with a focus on the ABL based on
observations \citep{hogstrom:BM1988,hogstrom:BM1996}
and numerical modelling \citep{spalart:PF2008, spalart:PF2009, ansorge:BM2014, ansorge:BM2019}.
%
\par
%
Attempts were also undertaken to obtain profiles of the wind speed: 
One approach is to match the inner and outer layer at a reference height as suggested by
\cite{etling:2002, emeis:2018}  (Sec. 21.10; Eq. 21.48); they choose the Prandtl-layer depth, 
a measure for the depth of the surface layer, to match the wind speed profiles. 
However, this requires external prescription of the veering at that height. 
A one-dimensional profile with constant veering is given by \citet[Sec. 3; Eq. 3.1-3.19]{emeis:m2007}. 
%%%

@@

homogeneous, stationary ABL over a flat surface and under neutral stratification. 
The description of velocity profiles for this strongly simplified problem is not only a canonical fluid-mechanical problem, but it 
also constitutes a well-described limit for theoretical exploration or higher-order approaches taking into account possible effects of stratification,

@@

(iv) Barotropic and baroclinic effects on the velocity profile require consideration of

@@

The f-plane approximation is applied, i.e. rotation acts on the horizontal components of velocity alone;
rotational effects on the vertical component of velocity and dynamical effects
due to latitudinal variation of the rate of rotation are neglected.

@@

correspond to an ensemble average. 

@@

  \underline{\tau}_\mathrm{sfc} = \left(\begin{array}{c} \tau_x\\ \tau_y\\\tau_z\end{array}\right) = - \rho\nu \left( \frac{\partial U}{\partial z} \hat{e}_x + \frac{\partial V}{\partial z} \hat{e}_y\right)

@@

  u_\star = \sqrt{ \frac{|| - \underline{\tau}_\mathrm{sfc} ||}{\rho} } 

@@

the only non-dimensional parameter that governs the system (in Ekman flow, the ratio of Rossby and Reynolds number 
are not independent as $Re_D=2\Lambda_\text{Ro} / D$).

@@

The outer normalization (with respect to the boundary-layer depth $\delta$ and velocity $u_\star$) is denoted 
by a superscript $'-'$, i.e. $z^-=z/\delta$.  

@@

The eddy viscosity plays a crucial part when modelling profiles and the vertical

@@

For the assumptions of horizontal homogeneity and incompressibility (which implies  $W=0$, i.e. no mean wall-normal velocity), it is

@@

    Buffer layer (i=10, $z^+\approx 10.5$),
    logarithmic layer (i=100, $z^+ \approx 150$), and
    outer layer (i=400, $z^+ \approx 1200$) of the case with $Re_\tau=2978$;

@@

\section{A Reynolds-number-independent velocity profile for the turbulent Ekman layer}

@@

boundary conditions at the surface but at a different height while maintaining the restriction 
to constant eddy viscosity. 

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Implications for surface-layer scaling} 
Eq.~(\ref{eqn:scaling_spanwise}) establishes a mixed scaling for the spanwise velocity in the viscous layer:
%
While it requires the vertical coordinate to be expressed in inner units, the velocity itself is normalized by the geostrophic wind, and becomes inversely proportional to the friction Reynolds number $\RE_\tau=\delta^+$ when considered at a fixed height.
%
In vicinity of the surface, such mixed scaling has already been identified for higher-order statistics in convective flows \citep{mellado:BM2016,li:JAS2018}, where large scales leave their signatures in vicinity of the surface.
%
It is important to note here that, while $V$ is a first-order statistic from a statistical perspective, the spanwise velocity is a higher-order correction term from the perspective of similarity theory and from the viewpoint of the channel-flow analogy that is routinely employed in the surface layer.
%
Further, this is consistent with the scaling for the velocity hodograph found in Eq.~(\ref{eqn:hodograph_scaling}) where the friction velocity also drops out.
%
\par
%
In the surface layer, there is not only a mixed scaling--as we had already identified in the viscous layer--, but we cannot find a universal function onto which the profiles of spanwise velocity collapse. 
%
This additional degree of freedom reflects the inner--outer matching problem for the spanwise velocity. 
%
Rather than giving a profile for this region, we resort here to a parametric description of the problem in terms of the function $f_{V,\mathrm{log}}$ determined by the parameters $a_\mathrm{log}$, $b_\mathrm{log}$, $c_\mathrm{log}$ which can be estimated based on the above scaling considerations for any Reynolds number.
%
We note that, once the parameter $a_\mathrm{log}$ is known, the parameters $b_\mathrm{log}$ and $c_\mathrm{log}$ can be estimated solely based on $f_{V,\mathrm{visc}}$, i.e. using the value $v_{10}$ and $d_{10}$ found for the viscous region of the flow.
%
For the range of Reynolds number relevant to geophysical problems ($10^4\lesssim \RE_D \lesssim 10^6$), the variation of $c_\mathrm{log}$ is, however, rather small.

@@

%
%\begin{figure}
%  \includegraphics[width=\textwidth]{mag_dir.eps} 
%  \caption{Total velocity and veering (in degrees) vs inner and outer height.
%    Dashed lines show DNS data, thick lines are from semi-empirical theory.} 
%\end{figure} 
%
\section{Conclusions}
%
We investigate the wind veer and fundamental scaling properties of the velocity profiles in Ekman flow. 
%
Based on scaling considerations and direct numerical simulation spanning on decade in external separation from $Re_\Lambda=1.25 \times 10^5$ to $Re_\Lambda=1.28\times 10^6$, we derive a closed formulation for both horizontal components of the velocity profile in a stationary, smooth Ekman layer.  
% 
This formulation is consistent with the DNS data and also yields reasonable results  at geophysical scale separation; the logarithmic law of the mean velocity is recovered with the well-known limits and deviations towards the surface and Ekman layer. 
%
The classic formulation of the Ekman layer, employing the surface boundary condition, is replaced by a modified solution that can be obtained by Ekman's system of governing equations, but using different boundary conditions that are more appropriate of the actual situation encountered in the planetary boundary layer. 
% 
The three parameters that characterize this boundary condition are estimated based on DNS data. 
% 
\par
% 
To quantify the spanwise velocity component consistently across the boundary layer, we derive a universal scaling of the spanwise velocity component in a shear-aligned reference frame. For the \emph{inner layer}, we find the mixed scaling  
%
\begin{equation} 
  V^{\alpha_\star}/G = \frac{1}{\delta^+} f(z^+)
\end{equation}
%
that is, the spanwise velocity normalized by the outer velocity scale is a universal function of the inner height and Friction Reynolds number $Re_\tau = \delta^+$. 
% 
This scaling is derived here based on scaling considerations, and it is in excellent agreement with the DNS data available.   
% 
In the outer layer, the spanwise velocity follows outer scaling, consistently with the Ekman model discussed above (Sec.~\ref{sec:ekman}).
% 
\par
% 
Our results suggest that there is no lower limit of the turning. 
% 
Hence---despite its very large scale separation / huge Reynolds number---the ABL is not 
in the 'limit' of high Reynolds number from the perspective of wind veer, but always in a regime where 
changes in $\RE$ impact on the vertical partitioning of rotation.

@@

\appendix
\section{Appendices}
\subsection{Laminar Ekman solution with consideration of inner layer}
\label{app:ekman_solution}
The following Ekman system is obtained by averaging the Navier--Stokes equations horizontally and over time and neglecting the turbulent transport terms (turbulence can then be considered via variation of the vertically constant eddy viscosity $\nu$; for a height-dependent eddy-viscosity model, the reader is referred to \cite{basu:BM2023}):
\begin{subequations}
\begin{eqnarray}
  \left(\begin{matrix}
    \p_t U\\
    \p_t V  
  \end{matrix}\right)&=&\left(\begin{matrix}
     fV &+& \nu \p_z^2 U\\ 
    -f(U-G) &+& \nu \p_z^2 V
  \end{matrix}\right)\\ 
  \Rightarrow \partial_t (U+iV) &=& f(V-i(U-G))) + \nu \p_z^2(U+iV)
\end{eqnarray}
In stationary conditions, this system is solved by
\begin{eqnarray}
  \hat{u}(z) &=& U_{\infty} + e^{-\gamma z} \left[A \cos\gamma z + B \sin\gamma z\right]   \\
  \hat{v}(z) &=& V_{\infty} + e^{-\gamma z} \left[-A \sin\gamma z + B \cos\gamma z\right]
\end{eqnarray} 
\end{subequations}
where the constants $U_\infty$, $V_\infty$ set the top boundary condition and $A$ and $B$ set the bottom boundary condition. 
%
The most common boundary condition for a surface Ekman layer is $A=U_{\infty}=G$, $B=0$, and $V_{\infty}=0$.
%
The lower boundary condition, however, neglects the existence of the surface layer, and it appears reasonable to define
$A=c G$ where $c<1$ is a constant that incorporates the increased shear in the surface layer.
%
Given a 'matching height' $z_{match}$ and normalized matching height $\xi=\gamma z_\text{match}$ in the upper part of the inner layer, we can match the Ekman profile
to the inner layer by letting
%
\begin{subequations}
\begin{eqnarray}
  \begin{matrix} 
    u(z_\text{match}) \equiv u_\text{match} &=& U_{\infty} + e^{-\gzm} \left [ A \cos\gzm + B\sin\gzm \right] \\
    v(z_\text{match}) \equiv v_\text{match} &=& V_{\infty} + e^{-\gzm} \left [-A \sin\gzm + B\cos\gzm \right]
  \end{matrix} 
\end{eqnarray}
\begin{eqnarray}
  \Rightarrow \left(\begin{matrix}
    u_\text{match}-U_{\infty} \\ 
    v_\text{match}-V_{\infty}
  \end{matrix}\right) = e^{-\gzm}\left(\begin{matrix}A \\ B \end{matrix}\right)\left(\begin{matrix}
    \cos\gzm &+ \sin\gzm \\ -\sin\gzm &+ \cos\gzm
  \end{matrix} \right) \\
\end{eqnarray}
\end{subequations}
Matching the profile at $\xi=0$, one obtains
$A = \Delta u_\text{match} $ and $B=-\Delta w_\text{match} $; and when the direction $Ox$ is aligned
with the geostrophic wind, we obtain the textbook-case $A=|\mathbf{G}|$ and $B=0$.

@@

\vspace{1em}
\noindent\textbf{Acknowledgements} We thank Sukanta Basu and two anonymous reviewers for their valuable reviews of this manuscript. Simulations were carried out on the German supercomputing systems 
\texttt{hawk} at H\"ocstleistungsrechenzentrum Stuttgart (Bundesprojekt 44187) and \texttt{juwels} at Gauss Centre for supercomputing, 
For\-schungs\-zen\-trum J\"ulich (Projects \texttt{hku24} and \texttt{stadit}). 
We extend our thanks to Sally Issa for valuable feedback on the manuscript.

@@

\bibitem[{Basu and Holtslag(2023)}]{basu:BM2023}
Basu S, Holtslag AAM (2023) A novel approach for deriving the stable boundary layer height and eddy viscosity concepts from the Ekman equations. Boundary-Layer Meteorol (187):105--115, \doi{}
